Synthesis of production rules for glycans

### 2020-10-14 papers/sugar-synth/algo.tex
many changes

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Here, we are considering only single compartment case.
The generalizations are discussed in section~\ref{sec:variations}.
Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
At a high level, our method is iterative.
The method first synthesizes a set of rules $R$ that can produce the molecules in input set $\mu$, i.e., $\mu \subseteq P(S,R)$.
However, there is also an additional requirement that a molecule
that is not in $\mu$ is not producible by the synthesized rules $R$.
Therefore, the method looks for the producible molecules that are not in $\mu$, called
{\em counterexample molecule}.
If there is no such molecule, our method terminates and reports
the synthesized rules.
If there is a molecule in $P(S,R) - \mu$, we add constraints that says
that look for rules
that do not generate
the molecule and go for another iteration.
Now let us look at the details of the method.

@@

The method assumes that the input set $\mu$ is finite, which a
natural assumption.
Even if a set of rules can produce
an unboundedly large number of molecules.
No real-life biology will exhibit such a set in a cell.
Since we are synthesizing production rules, our method
needs to bound the search space.
The method also takes two numbers as input:
$d$ is the maximum height of the learned rules
and 
$n$ is the maximum number of the learned rules.
The parameters eliminate potential
unbounded search space.
If the method fails to find production rules,
the user can always call the method with larger parameters.
However, we do not expect to find a large number of or sized rules
because chemical reactions do not usually have many conditional
on the precursors and do not apply large changes in a single step.
First, the method initializes $S$ with the set of monomers occurring
in $\mu$
and sets $w$ to be the maximum arity of any monomer in $S$.

We use templates to model the search space of rules.
A template is also a tree that has a certain depth and
the internal nodes of the tree have the same number of children.
Two variables label each node of the template.
One variable is for choosing the sugar at the node and the other is for describing
the `situation' of the node.
The first variable at the node encodes choices of
sugar monomers or their absence.
The domain of the variables is $S \union \{\bot\}$.
Let $SVars$ be the set of uncountable variables with the domain.
We will use the pool of $SVars$ to add variables to the templates.

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations of the nodes in a rule.
The first situation is when a node position may be in the expanding part.
The nodes in the dark gray area are the expanding nodes.
The second situation is when a node position may not be present in the rule.
The dashed area includes absent nodes
from the full tree.
Among the matching nodes, we make two cases, namely if the expanding
nodes are the descendants of the matching node or not. 
The third situation is when a node position is in the matching part
and has expanding decadents.
The solid path from $v_0$ to the root $v_e$ of the expanding part includes
the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is when a node position is in the matching part and
has no expanding decedents.
The nodes in the rest of the light gray area are in the fourth situation.
Another variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
We will use the pool of $KVars$ to add variables to the templates.
Formally, the templates are defined as follows.

\begin{figure}[t]
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(2mm,-6mm)$)
          ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}

  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
\end{figure}

@@

Let us consider the body of the loop in detail.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At lines~\ref{line:encodep-ans-match} and~\ref{line:encodep-ans-order}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
declared $MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The for loop at \ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
$\textsc{MatchTree}$ is also called with a third parameter
$mark$ and a Boolean value as the fourth parameter.
We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get the node name $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
After the while loop in $\textsc{EncodeP}$, $v_r$ is the
candidate node for the root of expanding nodes.
At line \ref{line:encodep-vr-expand}, we declare that $v_r$
is an $Expand$ node.
Afterword, we match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$ again at line~\ref{line:encodep-mtree-expand}.
At line~\ref{line:mcut}, we also call to match the cut pattern at
the subtree of $v$ with the template rule.

>>>>

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Here, we are considering only single compartment case.
The generalizations are discussed in section~\ref{sec:variations}.
Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.

% At a high level, our method is iterative.
% The method first synthesizes a set of rules $R$ that can produce the molecules in input set $\mu$, i.e., $\mu \subseteq P(S,R)$.
% However, there is also an additional requirement that a molecule
% that is not in $\mu$ is not producible by the synthesized rules $R$.
% Therefore, the method looks for the producible molecules that are not in $\mu$, called
% {\em counterexample molecule}.
% If there is no such molecule, our method terminates and reports
% the synthesized rules.
% If there is a molecule in $P(S,R) - \mu$, we add constraints that says
% that look for rules
% that do not generate
% the molecule and go for another iteration.
% Now let us look at the details of the method.

@@

The method assumes that the input set $\mu$ is finite.
Even if a set of rules can produce
an unbounded number of molecules.
No biology will exhibit such a set in a cell.
Our method
bounds the search space of production rules.
The method also takes two numbers as input:
$d$ is the maximum height of the learned rules
and 
$n$ is the maximum number of them.
% The parameters eliminate potential
% unbounded search space.
If the method fails to find production rules,
the user can always call the method with larger parameters.
% However, we do not expect to find a large number of or sized rules
% because chemical reactions do not usually have many conditional
% on the precursors and do not apply large changes in a single step.
First, the method initializes $S$ with the set of monomers occurring
in $\mu$
and sets $w$ to be the maximum arity of any monomer in $S$.

We use templates to model the search space of rules.
A template is also a tree that has a depth and
the internal nodes of the tree have the same number of children.
Two variables label each node of the template.
One variable is for choosing the sugar at the node and the other is for describing
the `situation' of the node.
% The first variable at the node encodes choices of
% sugar monomers or their absence.
The domain of the first variables is $S \union \{\bot\}$.
Let $SVars$ be the set of uncountable variables with the domain.
We will use the pool of $SVars$ to add variables to the templates.

\begin{wrapfigure}{r}{0.37\textwidth}
    \vspace{-6mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position may be in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position may not be present in the rule.
The dashed area includes absent nodes
from the full tree.
Among the matching nodes, we make two cases, namely if the expanding
nodes are the descendants of the matching node or not. 
The third situation is when a node position is in the matching part
and has expanding decadents.
The solid path from $v_0$ to the root $v_e$ of the expanding part includes
the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is when a node position is in the matching part and
has no expanding decedents.
The nodes in the rest of the light gray area are in the fourth situation.
Another variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
We will use the pool of $KVars$ to add variables to the templates.
Formally, the templates are defined as follows.

@@

Let us consider the body of the loop in detail.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
declared $MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The for loop at \ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
$\textsc{MatchTree}$ is also called with a third parameter
$mark$ and a Boolean value as the fourth parameter.
We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get the node name $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
After the while loop in $\textsc{EncodeP}$, $v_r$ is the
candidate node for the root of expanding nodes.
At line \ref{line:encodep-vr-expand}, we declare that $v_r$
is an $Expand$ node.
Afterword, we match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$ again at line~\ref{line:encodep-mtree-expand}.
At line~\ref{line:mcut}, we also call to match the cut pattern at
the subtree of $v$ with the template rule.

### 2020-10-14 papers/sugar-synth/bio.tex
many changes

\begin{figure}[t]
  \centering
  \begin{minipage}{0.54\linewidth}
    \includegraphics[width=0.9\linewidth]{gfig1.png}    
  \end{minipage}
  \begin{minipage}{0.44\linewidth}
    \caption{Biological details of glycan production. There are many types of sugar monomer building blocks (each represented by standard colored shapes); for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
    \label{fig:glycan-rule}
  \end{minipage}
\vspace{-9mm}
\end{figure}

>>>>

\begin{figure}[t]
  \centering
  \begin{minipage}{0.54\linewidth}
    \includegraphics[width=0.9\linewidth]{gfig1.png}    
  \end{minipage}
  \begin{minipage}{0.44\linewidth}
    \caption{Biological details of glycan production. There are many types of sugar monomer building blocks; %(each represented by standard colored shapes);
      for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
    \label{fig:glycan-rule}
  \end{minipage}
\vspace{-9mm}
\end{figure}

### 2020-10-14 papers/sugar-synth/experiments.tex
many changes

\subsection{Implementation} 
We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool} for
analyzing the performance of the technique in practice.
%
The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
to discharge the generated satisfiability queries.
%
We have developed a specialized file format for giving input to~\ourtool.
%
The format allows one to declare monomers, list molecules over the monomers
in the SMTLIB like format,
and pass the input parameters of the method.

\subsection{Benchmark}

We have applied our tool on four sets of molecules form
respiratory mucins of a cystic fibrosis patient (D1),
human chorionic gonadotropin from a cancer cell line (D2), horse chorionic gonadotropin (D3), and
hydra (D4)~\cite{Jaiman2018,hydra}.
The availability of clean data, where we are clear about the source and the collection
methodology used, limits our choices for the data.

\subsection{Results}
\input{tbl-results}

We have applied \ourtool~on the data set. For each data set, we choose several
parameter combinations to illustrate the relative performance of the tool.
If we did not give large enough parameters, then the tool fails to synthesize the rules.

>>>>

{\em Implementation:} 
We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool} for
analyzing the performance of the technique in practice.
%
The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
to discharge the generated satisfiability queries.
%
We have developed a specialized file format for giving input to~\ourtool.
%
The format allows one to declare monomers, list molecules over the monomers
in the SMTLIB like format,
and pass the input parameters of the method.

{\em Benchmark: }
We have applied our tool on four sets of molecules form
respiratory mucins of a cystic fibrosis patient (D1),
human chorionic gonadotropin from a cancer cell line (D2), horse chorionic gonadotropin (D3), and
hydra (D4)~\cite{Jaiman2018,hydra}.
The availability of clean data, where we are clear about the source and the collection
methodology used, limits our choices for the data.

\input{tbl-results}

{\em Results:} We have applied \ourtool~on the data set. For each data set, we choose several
parameter combinations to illustrate the relative performance of the tool.
If we did not give large enough parameters, then the tool fails to synthesize the rules.

### 2020-10-14 papers/sugar-synth/main.tex
many changes

\title{Automated inference of production rules for glycans}
\author{Ashutosh Gupta and Mukund Thattai and Ansuman Biswas}
%\institute{IITB and NCBS}

@@

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which these enzymes operate are not fully understood. In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans found in a cell. We have build a tool based on this method and have used it to infer rules based on published glycan data sets.
\end{abstract}

>>>>

\title{%Automated inference
  Synthesis of production rules for glycans}
\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
%\institute{IITB and NCBS}

@@

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which the enzymes operate are not fully understood. In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell. We have build a tool based on the method and applied it to infer rules based on published glycan data.
\end{abstract}

### 2020-10-14 papers/sugar-synth/model.tex
many changes

In this section, we present the formal model for the synthesis problem.
We model glycan molecules as labeled trees and production rules are also modeled using small pieces of trees. The glycan molecules are assembled by applying the production rules repeatedly.
Our synthesis problem reduces into finding the pieces of trees that represent the production rules.

Let $S$ be the set of sugar monomers that builds glycans,
the oligomer molecules. Each $s \in S$ is associated with arity $m$ 
(written $arity(s) = m$), {\em i.e.}, the maximum number of children for $s$. The children of the monomers are indexed, which means we can refer to the $k$th child of $s$ for some $k \leq arity(s)$.
In the monomers, they correspond to bonds at specific positions from where children are connected.
Now we define the glycan molecules as labeled trees. Now onward we refer to the glycans simply as molecules.

\begin{df}
A {\em molecule} $m = (V,M,C,v_0)$ is a labeled tree, where 
$V$ is a set of nodes in the tree,
$M : V \maps S$ maps nodes to their label, 
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes, and
$v_0 \in V$ is the root of the tree.
Furthermore, a molecule must satisfy the conditions of the arity of monomers.
Formally, let $M(v) = s$, if $C(v,n) = v'$ then $n \leq arity(s)$.
\end{df}

Let $m = (V,M,C,v_0)$ and $m' = (V',M',C',v_0')$ be molecules.
With some abuse of notation, we write $v \in m$ to denote $v \in V$.
For each $v \in m$, if $(v,n)$ is not in the domain of $C$, we write $C(v,n) = \bot$.
We assume that $C(v,0) = \bot$ and if $n > n' > 0$ and $C(v,n) \neq \bot$, then $C(v,n') \neq \bot$.
Let $NumberOfChildren(v)$ be equal to the number of $n$s such that $C(v,n) \neq \bot$. 
A node $v \in V$ is a {\em leaf} of $m$ if $C(v,n) = \bot$ for each $n$.
Let $depth(v)$ be the length of the path from $v_0$ to $v$.
A {\em branch} of $m$ is a path from $v_0$ to some leaf of $m$.
Let $height(m)$ be the length of the longest branch in $m$.
We define ancestor relation recursively as follows. Let $ancestor(m,v,0) = v$.
For for $d > 0$, let $ancestor(m,v,d) = ancestor(m,v', d-1)$ if $C(v',i) = v$ for some $i$.
Let recursively-defined predicate $Match(m,v,m',v')$ state that 
$v \in m$, $v' \in m'$, $s = M(v) = M(v')$, and $Match( m, C(v,n), m', C(v',n) )$ for each $n \leq arity(s)$ such that $C(v,n) \neq \bot$.
In other words, the subtree in $m$ rooted at $v$ is embedded in $m'$ at node $v'$.
Let $subtree(m)$ be the set of molecules such that
$m' = (\_,\_,\_,v_0') \in subtree(m) \lequiv Match(m,v_0,m',v_0')$.
Let us also define a utility to copy a subtree of a molecule into another molecule.
Let $Copy(m,v)$ return a molecule $m'' = (V'',M'',C'',v_0'')$ such that 
$V''$ is a set of fresh nodes, $Match(m,v,m'',v_0'')$, and $Match(m'',v_0'',m,v)$.
Both the $Match$ conditions say that the trees rooted at
$v$ and $v_0''$ are identical.

Now let us define the model of the production process of the molecules.
A production rule expands a molecule $m$ by attaching a new piece of tree at a node that has a vacant spot among its children if the surroundings of the
node satisfy some condition.
The rule is modeled as a tree that has two parts.
One part should already be there in $m$ and the other part will be appended to $m$.

\begin{df}
  A {\em production rule} $r = (V, M, C, v_0, v_e)$ is a labeled tree, where
  $V$ is a set of nodes in the tree,
  $M : V \maps S$ maps nodes to their label, 
  $C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
  $v_0 \in V$ is the root of the tree, and
  $v_e \in V$ is the root of expanding part of the rule.
\end{df}

If we {\em apply}, a rule $r$ on a molecule $m$, then it is extended at some node
$v \in m$. A copy of the descendants of $v_e$ will be be attached to
$v$ in $m$, and the rest of the nodes in the rule have to match $v$ and above.
We will refer to the descendants of $v_e$ as {\em expanding nodes}
and all the other nodes as {\em matching nodes}.

\begin{wrapfigure}{r}{0.37\textwidth}
  \vspace{-5mm}
    % \begin{minipage}{0.48\linewidth}
      \small
    % \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$B$};
    \node[sqloc, below left of=v0] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->] (v0) edge (v2);
  \end{tikzpicture}
  % \par
  % (a)
  % \end{minipage}
  % \begin{minipage}{0.48\linewidth}
    % \small
    % \center
  \quad
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$A$};
    \node[loc, below right of=v1] (v3) {$B$};
    \node[loc, below left of=v1] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->,dashed] (v1) edge (v2);
    \path[->] (v1) edge (v3);
  \end{tikzpicture}
  % \par
  \hfill (a) \hfill (b)\hfill\mbox{}
  % (b)
  % \end{minipage}
  \caption{(a) An example rule. (b) An application of the rule.}
  \label{fig:exrule}
  \vspace{-8mm}
\end{wrapfigure}
\paragraph{Example:} In Figure~\ref{fig:exrule}(a), we present a rule. That has
  two kinds of nodes.
  The rule adds the square node.
  Therefore, $v_e$ is the square node.
  The circular nodes %$A$ and its right child $B$
  are part of a pattern.
  The pattern must be present in the molecule to apply the rule.
  In Figure~\ref{fig:exrule}(b), we present an application of the rule.
  The solid tree with three nodes is the initial molecule.
  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
  Upon applying the rule, the rule adds a left child with label $A$ to the middle node.
  The rule is not applicable at the root $A$, since it has a right child $A$
  and the combination does not match the pattern.

@@

\subsection{Synthesis problem}
As discussed in section~\ref{sec:bio},
we observe a set of glycan molecules present in a cell.
%
However, we do not know the production rules to produce the molecules.
%
We will be developing a method to find the production rules.
%
Let us suppose we are given a set of molecules $\mu$ produced by some
unknown set of production rules.
%
The {\em synthesis problem} is to find a set $R$ of production rules
such that $\mu = P(S,R)$,
where $S$ is the set of monomers.

%
We can define a more general form of the synthesis problem.
Let us suppose we are also given $k$ compartments with unknown rules.
The goal of synthesis is to find  $R_1,...,R_k$ such that
$\mu = P(S, R_1,..,R_k)$.
We may relax the requirement of exactly producing $\mu$.
We may say that molecules in $\mu$ have to be produced, but we are ok
if subtrees of the molecules of $\mu$ are also produced.
Formally, we may weaken the requirement to $\mu \subseteq P(Q, R_1,..,R_k) \subseteq subtree(\mu)$.

The above are a few simplified versions of the biological problem.
We may generalize the problem further where
%rules are partitioned and the partitions are applied in phases one after another,
rules are not applied exhaustively due to time constraints,
the given $\mu$ is finite and may not be exhaustive,
and we only know the weights of the parts of molecules in $\mu$.
We will first present a method for solving a simplified version of the problem.
However, our tool handles some of the above variations.
We will discuss the variations in section~\ref{sec:variations}.

>>>>

In this section, we present the formal model for the synthesis problem.
We model glycan molecules and production rules as labeled trees. The glycan molecules are assembled by applying the production rules repeatedly.
Our synthesis problem reduces into finding the pieces of trees that represent the production rules.

Let $S$ be the set of sugar monomers that builds glycans,
the oligomer molecules. Each $s \in S$ is associated with arity $m$ 
(written $arity(s) = m$).
% , {\em i.e.}, the maximum number of children for $s$.
The children of the monomers are indexed. We refer to the $k$th child of $s$ for some $k \leq arity(s)$.
They correspond to bonds at specific positions in the monomers where children are connected.
Now we define the glycan molecules as labeled trees. Now onward we refer to the glycans simply as molecules.

\begin{df}
A {\em molecule} $m = (V,M,C,v_0)$ is a labeled tree, where 
$V$ is a set of nodes in the tree,
$M : V \maps S$ maps nodes to their label, 
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes, and
$v_0 \in V$ is the root of $m$.
A molecule must respect the arity of monomers, i.e., if $M(v) = s$ and $C(v,n) = v'$ then $n \leq arity(s)$.
\end{df}

Let $m = (V,M,C,v_0)$ and $m' = (V',M',C',v_0')$ be molecules.
With some abuse of notation, we write $v \in m$ to denote $v \in V$.
For each $v \in m$, if $(v,n)$ is not in the domain of $C$, we write $C(v,n) = \bot$.
We assume that $C(v,0) = \bot$.
%and if $n > n' > 0$ and $C(v,n) \neq \bot$, then $C(v,n') \neq \bot$.
Let $NumberOfChildren(v)$ be equal to the number of $n$s such that $C(v,n) \neq \bot$. 
A node $v \in V$ is a {\em leaf} of $m$ if $C(v,n) = \bot$ for each $n$.
Let $depth(v)$ be the length of the path from $v_0$ to $v$.
A {\em branch} of $m$ is a path from $v_0$ to some leaf of $m$.
Let $height(m)$ be the length of the longest branch in $m$.
We define ancestor relation recursively as follows. Let $ancestor(m,v,0) = v$.
For for $d > 0$, let $ancestor(m,v,d) = ancestor(m,v', d-1)$ if $C(v',i) = v$ for some $i$.
Let recursively-defined predicate $Match(m,v,m',v')$ state that 
$v \in m$, $v' \in m'$, $s = M(v) = M(v')$, and $Match( m, C(v,n), m', C(v',n) )$ for each $n \leq arity(s)$ such that $C(v,n) \neq \bot$.
In other words, the subtree in $m$ rooted at $v$ is embedded in $m'$ at node $v'$.
Let $subtree(m)$ be the set of molecules such that
$m' = (\_,\_,\_,v_0') \in subtree(m) \lequiv Match(m,v_0,m',v_0')$.
Let us also define a utility to copy a subtree of a molecule into another molecule.
Let $Copy(m,v)$ return a molecule $m'' = (V'',M'',C'',v_0'')$ such that 
$V''$ is a set of fresh nodes, $Match(m,v,m'',v_0'')$, and $Match(m'',v_0'',m,v)$.
Both the $Match$ conditions say that the trees rooted at
$v$ and $v_0''$ are identical.

Now we define the model of the production process of the molecules.
A production rule expands a molecule $m$ by attaching a new piece of tree at a node that has a vacant spot among its children if the surroundings of the
node satisfy some condition.
The rule is modeled as a tree that has two parts.
One part should already be there in $m$ and the other part will be appended to $m$.

\begin{df}
  A {\em production rule} $r = (V, M, C, v_0, v_e)$ is a labeled tree, where
  $V$ is a set of nodes,
  $M : V \maps S$ maps nodes to labels, 
  $C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
  $v_0 \in V$ is the root, and
  $v_e \in V$ is the root of expanding part of the rule.
\end{df}

If we {\em apply}, a rule $r$ on a molecule $m$, then it is extended at some node
$v \in m$. A copy of the descendants of $v_e$ will be attached to
$v$ in $m$, and the rest of the nodes in the rule have to match $v$ and above.
We call the descendants of $v_e$ as {\em expanding nodes}
and all the other nodes as {\em matching nodes}.

\begin{wrapfigure}{r}{0.37\textwidth}
  \vspace{-12mm}
    % \begin{minipage}{0.48\linewidth}
      \small
    % \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$B$};
    \node[sqloc, below left of=v0] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->] (v0) edge (v2);
  \end{tikzpicture}
  % \par
  % (a)
  % \end{minipage}
  % \begin{minipage}{0.48\linewidth}
    % \small
    % \center
  \quad
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$A$};
    \node[loc, below right of=v1] (v3) {$B$};
    \node[loc, below left of=v1] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->,dashed] (v1) edge (v2);
    \path[->] (v1) edge (v3);
  \end{tikzpicture}
  % \par
  \hfill (a) \hfill (b)\hfill\mbox{}
  % (b)
  % \end{minipage}
  \caption{(a) A rule. (b) An application of the rule.}
  \label{fig:exrule}
  \vspace{-8mm}
\end{wrapfigure}
\paragraph{Example:} In Figure~\ref{fig:exrule}(a), we present a rule. It has
  two kinds of nodes.
  The rule adds the square node ($v_e$).
  The circular nodes %$A$ and its right child $B$
  are the pattern, which must be present in the molecule to apply the rule.
  In Figure~\ref{fig:exrule}(b), we present an application of the rule.
  The solid tree with three nodes is the initial molecule.
  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
  %Upon applying the rule,
  The rule adds a left child with label $A$ to the middle node.
  The rule is not applicable at the root $A$ due to pattern mismatch.
  % , since it has a right child $A$
  % and the combination does not match the pattern.

@@

\subsection{Synthesis problem}
We observe a set of glycan molecules $\mu$ present in a cell.
%
However, we do not know the production rules to produce the molecules.
%
We will be developing a method to find the production rules.
%
The {\em synthesis problem} is to find a set $R$ of production rules
such that $\mu = P(S,R)$,
where $S$ is the set of monomers.

% %
% We can define a more general form of the synthesis problem.
% Let us suppose we are also given $k$ compartments with unknown rules.
% The goal of synthesis is to find  $R_1,...,R_k$ such that
% $\mu = P(S, R_1,..,R_k)$.
% We may relax the requirement of exactly producing $\mu$.
% We may say that molecules in $\mu$ have to be produced, but we are ok
% if subtrees of the molecules of $\mu$ are also produced.
% Formally, we may weaken the requirement to $\mu \subseteq P(Q, R_1,..,R_k) \subseteq subtree(\mu)$.

% The above are a few simplified versions of the biological problem.
% We may generalize the problem further where
% %rules are partitioned and the partitions are applied in phases one after another,
% rules are not applied exhaustively due to time constraints,
% the given $\mu$ is finite and may not be exhaustive,
% and we only know the weights of the parts of molecules in $\mu$.
% We will first present a method for solving a simplified version of the problem.
% However, our tool handles some of the above variations.
% We will discuss the variations in section~\ref{sec:variations}.
